### 2025-08-30 chapters/discussion.tex
review complete

This chapter transitions from measurement to meaning: it connects the observed data to the architectural principles underpinning enclave execution, therebyinforming both future research and practical deployment of secure computing systems based on open RISC-V architectures.

@@

However, workloads with more intensive memory access or intricate memory patterns, such as Kyber (a post-quantum cryptographic algorithm), experience much higher overhead. The Kyber algorithm, which involves significant pointer dereferencing and array indexing, incurs a significant penalty as every memory access is subject to additional security checks and encryption layers. This is compounded by the latency-sensitive nature of cryptographic workloads, where even small delays can result in significant performance degradation. The nearly threefold slowdown in enclave execution compared to native execution is primarily due to the added latency from memory encryption and integrity checks, which are inherently more expensive for memory-intensive algorithms.

@@

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1, transform shape]
\begin{axis}[
    ybar,
    bar width=10pt,
    width=0.9\linewidth, height=6cm,
    ylabel={PMP entries required},
    symbolic x coords={128\,KB aligned,192\,KB misaligned,300\,KB misaligned,384\,KB misaligned},
    xtick=data,
    x tick label style={rotate=20, anchor=east},
    ymajorgrids, grid style=dashed,
    enlarge x limits=0.15,
    legend style={at={(0.5,-0.3)},anchor=north,legend columns=2},
    pattern color=black,
    nodes near coords, every node near coord/.append style={font=\scriptsize, rotate=90, anchor=west}
]
\addplot[pattern=north east lines] coordinates {
(128\,KB aligned,1)
(192\,KB misaligned,2)
(300\,KB misaligned,2)
(384\,KB misaligned,1)
};
\addplot[pattern=crosshatch] coordinates {
(128\,KB aligned,2)
(192\,KB misaligned,3)
(300\,KB misaligned,3)
(384\,KB misaligned,2)
};
\legend{Contiguous mapping,Fragmented mapping}
\end{axis}
\end{tikzpicture}
\caption{Fragmentation and misalignment increase PMP entry usage, showing how non-ideal alignment leads to more PMP entries per enclave.}
\label{fig:pmp-fragmentation-entries}
\end{figure}

\begin{lstlisting}[language=Python, caption={Multiplexed enclave task dispatcher.}, label={lst:enclave-mux}]
def enclave_dispatcher(task_id, input_data):
    if task_id == 1:
        return process_crypto(input_data)
    elif task_id == 2:
        return handle_key_store(input_data)
    elif task_id == 3:
        return run_diagnostics(input_data)
    else:
        raise ValueError("Unknown task")
\end{lstlisting}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1, transform shape,font=\small]
% Memory bar
\draw[thick] (0,0) rectangle (12,1);
\node[anchor=south] at (0,1.1) {Base};
\node[anchor=south] at (12,1.1) {Top};

@@

\node[anchor=west] at (0,-1.5) {\footnotesize Hatched = padding/waste due to power-of-two alignment};
\end{tikzpicture}
\caption{Fragmentation overhead caused by mapping non-contiguous memory regions to NAPOT-aligned blocks. The hatched areas represent wasted padding required for alignment.}
\label{fig:napot-fragmentation}
\end{figure}

@@

A fundamental consideration when configuring RISC-V enclaves involves a holistic understanding of the interplay between hardware resources, isolation mechanisms, and workload characteristics. As discussed in Section~\ref{sec:synthesis} of Chapter~\ref{chap:discussion}, the Physical Memory Protection (PMP) mechanism stands at the heart of this relationship, providing the essential hardware-enforced boundary that isolates enclave memory from the untrusted environment. However, PMP’s inherent design—specifically the limited number of entries available and the strict Naturally Aligned Power-of-Two (NAPOT) encoding format for memory regions—imposes a critical scalability constraint that directly influences enclave instantiation and concurrency. Consequently, configuring enclave memory regions to maximize contiguity and alignment is paramount; doing so minimizes PMP entry consumption, reduces fragmentation, and alleviates potential bottlenecks in enclave creation and management. Developers are advised to meticulously design memory layouts that conform to PMP’s alignment constraints, thereby ensuring that enclave memory mappings are as compact and contiguous as possible (see Figure~\ref{fig:pmp-fragmentation-group}).

@@

Given the PMP entry limitations and the observed overheads of enclave context switches, a key deployment recommendation is to limit the number of simultaneously active enclaves to avoid PMP exhaustion, which can lead to explicit enclave creation failures (see Section~\ref{sec:workload-sensitivity}, Chapter~\ref{chap:benchmarking}). Practical configurations should, therefore, consolidate workloads into fewer, more capable enclaves when possible—a strategy known as enclave multiplexing. This approach not only conserves precious PMP entries but also reduces synchronization overhead and inter-enclave communication latency. Software-level techniques, including dynamic PMP entry management and runtime reconfiguration, can further optimize resource utilization by adapting PMP assignments based on workload activity and priority, though these require sophisticated coordination to maintain security invariants.

>>>>

This chapter transitions from measurement to meaning: it connects the observed data to the architectural principles underpinning enclave execution, thereby informing both future research and practical deployment of secure computing systems based on open RISC-V architectures.

@@

However, workloads with more intensive memory access or intricate memory patterns, such as Kyber (a post-quantum cryptographic algorithm), experience much higher overhead. Kyber involves frequent pointer dereferencing and array indexing, where each operation depends on the completion of the previous memory access. Within an enclave, every memory access triggers additional security checks, including encryption, integrity verification, and, in some cases, TLB flushes on context switches, adding fixed latency per operation.

Cryptographic workloads are particularly sensitive to such latency because the sequential, dependent nature of their computations prevents effective hiding of memory delays. Even small delays accumulate across thousands of operations, resulting in a substantial slowdown. Consequently, the nearly threefold slowdown observed for Kyber in enclave execution compared to native execution is primarily due to the added latency from memory encryption, integrity checks, and occasional TLB flushes, which disproportionately impact memory-intensive, latency-bound workloads.

@@

\begin{figure}[htbp]
\centering

% --- PMP entry usage chart ---
\begin{tikzpicture}[scale=1, transform shape]
\begin{axis}[
    ybar,
    bar width=10pt,
    width=0.9\linewidth, height=6cm,
    ylabel={PMP entries required},
    symbolic x coords={128\,KB aligned,192\,KB misaligned,300\,KB misaligned,384\,KB misaligned},
    xtick=data,
    x tick label style={rotate=20, anchor=east},
    ymajorgrids, grid style=dashed,
    enlarge x limits=0.15,
    legend style={at={(0.5,-0.4)},anchor=north,legend columns=2},
    pattern color=black,
    nodes near coords, every node near coord/.append style={font=\scriptsize, rotate=90, anchor=west}
]
\addplot[pattern=north east lines] coordinates {
(128\,KB aligned,1)
(192\,KB misaligned,2)
(300\,KB misaligned,2)
(384\,KB misaligned,1)
};
\addplot[pattern=crosshatch] coordinates {
(128\,KB aligned,2)
(192\,KB misaligned,3)
(300\,KB misaligned,3)
(384\,KB misaligned,2)
};
\legend{Contiguous mapping,Fragmented mapping}
\end{axis}
\end{tikzpicture}

\vspace{1em}

% --- NAPOT fragmentation diagram ---
\begin{tikzpicture}[scale=1, transform shape,font=\small]
% Memory bar
\draw[thick] (0,0) rectangle (12,1);
\node[anchor=south] at (0,1.1) {Base};
\node[anchor=south] at (12,1.1) {Top};

@@

\node[anchor=west] at (0,-1.5) {\footnotesize Hatched = padding/waste due to power-of-two alignment};
\end{tikzpicture}

\caption{Top: PMP entry usage increases with fragmentation and misalignment. Bottom: Fragmentation overhead caused by mapping non-contiguous memory regions to NAPOT-aligned blocks, with hatched areas representing wasted padding.}
\label{fig:pmp-and-napot-fragmentation}
\end{figure}

To address overhead and scalability challenges, \emph{enclave multiplexing} offers a promising software-level strategy by consolidating multiple lightweight workloads into a single enclave. This reduces the number of PMP entries required by reusing them across tasks, optimizing resource usage and improving scalability. An example of a multiplexed enclave task dispatcher is shown in Listing~\ref{lst:enclave-mux}.
\vspace{8em}
\begin{lstlisting}[language=Python, caption={Multiplexed enclave task dispatcher.}, label={lst:enclave-mux}]
def enclave_dispatcher(task_id, input_data):
    if task_id == 1:
        return process_crypto(input_data)
    elif task_id == 2:
        return handle_key_store(input_data)
    elif task_id == 3:
        return run_diagnostics(input_data)
    else:
        raise ValueError("Unknown task")
\end{lstlisting}

% \begin{figure}[htbp]
% \centering
% \begin{tikzpicture}[scale=1, transform shape,font=\small]
% % Memory bar
% \draw[thick] (0,0) rectangle (12,1);
% \node[anchor=south] at (0,1.1) {Base};
% \node[anchor=south] at (12,1.1) {Top};

% % Needed regions (non-contiguous)
% \draw[fill=gray!30] (0.5,0.1) rectangle (2.2,0.9) node[pos=.5] {\footnotesize R1};
% \draw[fill=gray!30] (3.8,0.1) rectangle (4.6,0.9) node[pos=.5] {\footnotesize R2};
% \draw[fill=gray!30] (7.0,0.1) rectangle (9.1,0.9) node[pos=.5] {\footnotesize R3};

% % NAPOT allocations (outer boxes) and padding (hatched)
% \draw[thick] (0.25,-0.3) rectangle (2.5,-1.1) node[pos=.5] {\footnotesize NAPOT fit R1};
% \draw[pattern=north east lines,pattern color=black] (0.25,-0.3) rectangle (0.5,-1.1);
% \draw[pattern=north east lines,pattern color=black] (2.2,-0.3) rectangle (2.5,-1.1);

% \draw[thick] (3.5,-0.3) rectangle (5.1,-1.1) node[pos=.5] {\footnotesize NAPOT fit R2};
% \draw[pattern=north east lines,pattern color=black] (3.5,-0.3) rectangle (3.8,-1.1);
% \draw[pattern=north east lines,pattern color=black] (4.6,-0.3) rectangle (5.1,-1.1);

% \draw[thick] (6.7,-0.3) rectangle (9.3,-1.1) node[pos=.5] {\footnotesize NAPOT fit R3};
% \draw[pattern=north east lines,pattern color=black] (6.7,-0.3) rectangle (7.0,-1.1);
% \draw[pattern=north east lines,pattern color=black] (9.1,-0.3) rectangle (9.3,-1.1);

% \node[anchor=west] at (0,-1.5) {\footnotesize Hatched = padding/waste due to power-of-two alignment};
% \end{tikzpicture}
% \caption{Fragmentation overhead caused by mapping non-contiguous memory regions to NAPOT-aligned blocks. The hatched areas represent wasted padding required for alignment.}
% \label{fig:napot-fragmentation}
% \end{figure}

@@

A fundamental consideration when configuring RISC-V enclaves involves a holistic understanding of the interplay between hardware resources, isolation mechanisms, and workload characteristics. As discussed in Section~\ref{sec:synthesis} of Chapter~\ref{chap:discussion}, the Physical Memory Protection (PMP) mechanism stands at the heart of this relationship, providing the essential hardware-enforced boundary that isolates enclave memory from the untrusted environment. However, PMP’s inherent design—specifically the limited number of entries available and the strict Naturally Aligned Power-of-Two (NAPOT) encoding format for memory regions—imposes a critical scalability constraint that directly influences enclave instantiation and concurrency. Consequently, configuring enclave memory regions to maximize contiguity and alignment is paramount; doing so minimizes PMP entry consumption, reduces fragmentation, and alleviates potential bottlenecks in enclave creation and management. Developers are advised to meticulously design memory layouts that conform to PMP’s alignment constraints, thereby ensuring that enclave memory mappings are as compact and contiguous as possible (see Figure~\ref{fig:pmp-and-napot-fragmentation}).

@@

Given the PMP entry limitations and the observed overheads of enclave context switches, a key deployment recommendation is to limit the number of simultaneously active enclaves to avoid PMP exhaustion, which can lead to explicit enclave creation failures (see Section~\ref{sec:workload-sensitivity}, Chapter~\ref{chap:benchmarking}). Practical configurations should, therefore, consolidate workloads into fewer, more capable enclaves when possible—a strategy known as enclave multiplexing \cite{Occlum2020}. This approach not only conserves precious PMP entries but also reduces synchronization overhead and inter-enclave communication latency. Software-level techniques, including dynamic PMP entry management and runtime reconfiguration, can further optimize resource utilization by adapting PMP assignments based on workload activity and priority, though these require sophisticated coordination to maintain security invariants.

### 2025-08-30 chapters/system-architecture.tex
review complete

To assess the performance implications of enclave-based isolation, each benchmark was implemented as a standalone enclave application (referred to as an \textit{eapp}), accompanied by a corresponding host application responsible for enclave lifecycle management. The host application, running in user space, handles enclave initialization, benchmark loading, invocation of enclave entry points, and inter-domain communication via edge calls. In contrast, the enclave application is fully isolated by the Keystone runtime and contains only the benchmark logic and associated data structures. This strict separation ensures that the same benchmark codebase can be executed both natively (without isolation) and securely (within an enclave), without requiring structural changes—thus enabling a fair and consistent comparative analysis.

@@

For kernel-level enclave debugging, QEMU’s built-in GDB server can be activated via the \texttt{KEYSTONE\_DEBUG} flag. Once the system boots, the Keystone kernel driver is loaded manually using the command in Listing~\ref{lst:modprobe}.

\begin{lstlisting}[language=bash, caption={Loading the Keystone kernel driver}, label={lst:modprobe}]
modprobe keystone-driver
\end{lstlisting}

@@

The specific functions evaluated are shown in Listing~\ref{lst:kyber_api}, grouped by security level:

\begin{lstlisting}[language=bash,caption={Kyber cryptographic API functions},label={lst:kyber_api}]
indcpa_keypair   # Key pair generation for IND-CPA variant
indcpa_enc       # Encryption (encapsulation) under IND-CPA assumptions
indcpa_dec       # Decryption (decapsulation) under IND-CPA assumptions

kyber_keypair    # Enhanced key pair generation supporting IND-CCA
kyber_encaps     # IND-CCA-compliant encapsulation
kyber_decaps     # IND-CCA-compliant decapsulation
\end{lstlisting}
Each of these operations was executed across the five representative system configurations defined in Section~\ref{sec:param-variation}, which span a broad range of memory capacities, core counts, and execution modes. To ensure measurement accuracy and reduce variability from transient system states, ten iterations were run per configuration with the first iteration discarded as a warm-up phase. Execution timing was obtained using the enclave’s internal high-resolution timer API, capturing secure-world runtime exclusively. In parallel, external system-level metrics — such as CPU usage, memory footprint, and context switching counts — were gathered from the host environment to provide contextual insights into system overhead and resource utilization.

@@

For reproducibility, all benchmarking scripts, system configurations, QEMU command-line options, Keystone runtime parameters, and cache plugin settings were version-controlled using Git. Raw benchmark outputs, kernel logs, and cache plugin trace files were archived and documented. A complete snapshot of the experimental infrastructure, along with scripts to regenerate results, is provided in Appendix~\ref{appendix:cache-plugin}. This archival effort ensures that all reported results can be independently verified and reproduced by future researchers or evaluators.

>>>>

To assess the performance implications of enclave-based isolation, each benchmark was implemented as a standalone enclave application (referred to as an \textit{eapp}), accompanied by a corresponding host application responsible for enclave lifecycle management. The host application, running in user space, handles enclave initialization, benchmark loading, invocation of enclave entry points, and inter-domain communication via edge calls, while the enclave application \texttt{(eapp)} also runs in user mode. In contrast, the enclave application is fully isolated by the Keystone runtime and contains only the benchmark logic and associated data structures. This strict separation ensures that the same benchmark codebase can be executed both natively (without isolation) and securely (within an enclave), without requiring structural changes—thus enabling a fair and consistent comparative analysis.

@@

For kernel-level enclave debugging, QEMU’s built-in GDB server can be activated via the \texttt{KEYSTONE\_DEBUG} flag. Once the system boots, the Keystone kernel driver is loaded manually using the command: \texttt{modprobe keystone-driver}.

@@

The specific functions evaluated are shown in Listing~\ref{lst:kyber_api}, grouped by security level:
\vspace{5em}
\begin{lstlisting}[language=bash,caption={Kyber cryptographic API functions},label={lst:kyber_api}]
indcpa_keypair   # Key pair generation for IND-CPA variant
indcpa_enc       # Encryption (encapsulation) under IND-CPA assumptions
indcpa_dec       # Decryption (decapsulation) under IND-CPA assumptions

kyber_keypair    # Enhanced key pair generation supporting IND-CCA
kyber_encaps     # IND-CCA-compliant encapsulation
kyber_decaps     # IND-CCA-compliant decapsulation
\end{lstlisting}
Each of these operations was executed across the five representative system configurations defined in Section~\ref{sec:param-variation}, which span a broad range of memory capacities, core counts, and execution modes. To ensure measurement accuracy and reduce variability from transient system states, ten iterations were run per configuration with the first iteration discarded as a warm-up phase. Execution timing was obtained using the enclave’s internal high-resolution timer API, capturing only secure-world runtime and thus remaining isolated from background activity in the Linux kernel. In parallel, external system-level metrics — including CPU usage, memory footprint, and context switching counts — were gathered from the host environment, reflecting overall system overhead and resource utilization influenced by the kernel and other processes.

@@

For reproducibility, all benchmarking scripts, system configurations, QEMU command-line options, Keystone runtime parameters, and cache plugin settings were version-controlled using Git. Raw benchmark outputs, kernel logs, and cache plugin trace files were archived and documented. Complete snapshots of the experimental infrastructure, along with scripts to regenerate results, are publicly available\footnote{\url{https://github.com/coast-moonlight/keystone.git}}. This archival effort ensures that all reported results can be independently verified and reproduced by future researchers or evaluators.

### 2025-08-30 preamble.tex
review complete

\UzLThesisSetup{
Masterarbeit,
Verfasst              = {am}{Institut für Technische Informatik},
Titel auf Deutsch     = {
    Bewertung von RISC-V-Enklaven: Leistungsbenchmarks und bewährte Konfigurationspraktiken
}, 
Titel auf Englisch    = {
    Evaluating RISC-V Enclaves: Performance Benchmarks and Configuration Best Practices 
},
Autor                 = {Basil Ugbomoiko},
Betreuerin            = {Dr.-Ing. Saleh Mulhem},
Mit Unterstützung von = {Henry Strunck},
Studiengang           = {IT Security},
Datum                 = {31. August 2025},
Abstract = {This thesis evaluates the performance of Keystone enclaves, an open-source Trusted Execution Environment (TEE) designed for RISC-V platforms. Keystone enables secure computation by isolating sensitive workloads from the rest of the system. To assess its performance, the study benchmarks critical system parameters, including the number of CPU cores, cache size, memory allocation, and enclave configuration. Using industry-standard benchmarking tools such as lmbench and CoreMark, the research measures execution time, context switch overhead, memory latency, and CPU utilization across a range of hardware and software configurations. The experimental analysis identifies key performance bottlenecks that impact the efficiency of enclave execution. Based on these findings, the study presents configuration best practices and tuning recommendations tailored to various workload profiles, such as compute-bound, memory-intensive, and mixed applications. The insights gained from this evaluation contribute to optimizing the deployment of Keystone enclaves, making them more viable for real-world use cases that demand secure and efficient execution on RISC-V architectures.},
Zusammenfassung = {Diese Arbeit bewertet die Leistung von Keystone-Enklaven, einer Open-Source Trusted Execution Environment (TEE) für RISC-V-Plattformen. Keystone ermöglicht sichere Berechnungen, indem sensible Workloads vom restlichen System isoliert werden. Zur Leistungsbewertung werden zentrale Systemparameter wie die Anzahl der CPU-Kerne, die Cache-Größe, die Speicherzuweisung und die Konfiguration der Enklave untersucht. Mithilfe standardisierter Benchmarking-Tools wie lmbench und CoreMark werden Ausführungszeit, Kontextwechsel-Overhead, Speicherlatenz und CPU-Auslastung unter verschiedenen Hardware- und Softwarekonfigurationen gemessen. Die experimentelle Analyse identifiziert wichtige Leistungsengpässe, die die Effizienz der Enklaven beeinträchtigen. Auf Basis dieser Erkenntnisse werden Konfigurationsrichtlinien und Optimierungsempfehlungen vorgestellt, die auf unterschiedliche Workload-Profile zugeschnitten sind, z. B. rechenintensive, speicherintensive und gemischte Anwendungen. Die gewonnenen Erkenntnisse tragen dazu bei, den Einsatz von Keystone-Enklaven zu optimieren und ihre praktische Anwendbarkeit in sicherheitskritischen Bereichen auf RISC-V-Architekturen zu verbessern.},
Numerische Bibliographie
}

>>>>

\UzLThesisSetup{
Masterarbeit,
Logo-Dateiname        = {uzl-iti-logo-uzl.pdf},
Verfasst              = {am}{Institut für Technische Informatik},
Titel auf Deutsch     = {
    Bewertung von RISC-V-Enklaven: Leistungsbenchmarks und bewährte Konfigurationspraktiken
}, 
Titel auf Englisch    = {
    Evaluating RISC-V Enclaves: Performance Benchmarks and Configuration Best Practices 
},
Autor                 = {Basil Ugbomoiko},
Betreuerin            = {Dr.-Ing. Saleh Mulhem},
Mit Unterstützung von = {Henrik Strunck, M.Sc},
Studiengang           = {IT Security},
Datum                 = {01. September 2025},
Abstract = {
  This thesis evaluates the performance of Keystone enclaves, an open-source Trusted Execution Environment (TEE) for RISC-V platforms. Keystone isolates sensitive workloads to enable secure computation, and its performance is assessed across CPU cores, cache sizes, memory allocations, and enclave configurations. Using Dhrystone, CoreMark, and cryptographic workloads such as Kyber—a post-quantum key encapsulation mechanism—this work measures execution time, context switch overhead, memory latency, and CPU utilization. The results reveal critical bottlenecks and provide best-practice configuration guidelines for compute-bound, memory-intensive, mixed, and post-quantum workloads, supporting efficient and secure deployment of Keystone enclaves on RISC-V architectures.
},
Zusammenfassung = {
  Diese Arbeit bewertet die Leistung von Keystone-Enklaven, einer Open-Source Trusted Execution Environment (TEE) für RISC-V-Plattformen. Keystone isoliert sensible Workloads zur sicheren Berechnung, wobei Leistungseinflüsse durch CPU-Kerne, Cache-Größe, Speicherzuweisung und Enklavenkonfiguration untersucht werden. Mithilfe von Dhrystone, CoreMark und kryptografischen Workloads wie Kyber—einem postquantischen Schlüssel-Kapselungsverfahren—werden Ausführungszeit, Kontextwechsel-Overhead, Speicherlatenz und CPU-Auslastung gemessen. Die Ergebnisse zeigen zentrale Engpässe auf und liefern Konfigurationsrichtlinien für rechenintensive, speicherintensive, gemischte und postquantische Workloads, um Keystone-Enklaven effizient und sicher auf RISC-V-Architekturen einzusetzen.
},
Acknowledgements = {
    I would like to sincerely thank Daniel K. for his insightful guidance and clarifications, which added immense value to this work.
    I am also grateful to Tom K. for providing the initial log analysis script and for his valuable input on effective visualization approaches.
    My heartfelt thanks go to my parents for their patience, encouragement, and the opportunities they have provided.  
    I deeply appreciate my supervisors, Dr.-Ing. Saleh M. and Henrik S. (M.Sc), for their guidance, constructive feedback, and steadfast support throughout this project.  
    Finally, I extend my gratitude to my fellow Lübeckers and all others who contributed in various ways; their support, though too numerous to mention individually, was invaluable.
},
Numerische Bibliographie
}
